\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{tocloft}
\usepackage{listings}
\usepackage{verbatim}

\title{\textbf{Product Requirements Document (PRD)}\\
\textbf{AgeOfAgents}\\[0.5em]
\large The Ultimate Local-First RTS Command Center for AI Coding Agents}
\author{Dylan Kawalec (ChaoticBeautys) in collaboration with Grok AI}
\date{\today}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=cyan,
    pdftitle={AgeOfAgents PRD},
    pdfpagemode=FullScreen,
}

\begin{document}

\maketitle

\tableofcontents
\newpage

\section{Executive Summary \& Vision}

\textbf{AgeOfAgents} is a complete fork and re-imagining of the open-source \textbf{openage} engine[](https://github.com/SFTtech/openage) that transforms the classic \textit{Age of Empires II} real-time strategy (RTS) gameplay into the ultimate visual command center for orchestrating fleets of local AI coding agents.

Instead of managing peasants, knights, and castles, you will command **AI agents** (spawned from your own Ollama models and custom agent software) across a living, infinite battlefield where:
\begin{itemize}
    \item The **map = your entire massive monorepo + hundreds of internal repositories**.
    \item **Buildings = code modules, services, folders, and sub-repos**.
    \item **Units = AI coding agents** (Coder, Planner, Reviewer, Executor, Researcher, etc.).
    \item **Resources = compute, tokens, memory, and data**.
    \item **Tech tree = agent skill upgrades and new capabilities**.
    \item **Fog of war = unexplored parts of your protocol/architecture**.
\end{itemize}

We extract **every single idea** from AgentCraft (raw concepts from the April 2026 video https://www.youtube.com/watch?v=kR64LOqBBCU), Arcane Agents, VibeCraft, and GastownRTS, then replace the original AoE2 war logic with production-grade AI orchestration.

\textbf{Core Promise}: Run everything 100\% locally. Point AgeOfAgents at your giant monorepo root. Spawn, monitor, and command hundreds of Ollama-powered agents in true RTS style. The battlefield visually represents your entire codebase architecture in real time.

This PRD is written to be **production-ready for AI coding agents** (Claude Code, Cursor, Ollama-based agents, etc.) to implement the entire system end-to-end.

\textbf{Project Name}: \textbf{AgeOfAgents} (or ``Age of Agents'')

\textbf{Base Repository to Fork}: Your fork \href{https://github.com/DylanCkawalec/openage}{https://github.com/DylanCkawalec/openage} (which is itself a fork of the upstream \href{https://github.com/SFTtech/openage}{SFTtech/openage}).

\textbf{License}: Remain GPLv3 (inherited from openage).

\newpage
\section{Project Inspirations \& References}

All raw ideas are consolidated from the following open-source projects and the AgentCraft video:

\begin{itemize}
    \item \textbf{AgentCraft} – Raw ideas only (no npx usage) \\
      Video: \href{https://www.youtube.com/watch?v=kR64LOqBBCU}{https://www.youtube.com/watch?v=kR64LOqBBCU} \\
      Core concepts: RTS UI, file-system-as-map, buildings-as-functionality, campaign orchestrator, container isolation, review bundles, heat maps, quest system.
    \item \textbf{Arcane Agents} – \href{https://github.com/thomasrice/arcane-agents}{https://github.com/thomasrice/arcane-agents} \\
      Any CLI runtime (Ollama), live tmux terminal streaming, auto status detection, project auto-discovery, keyboard-first hotkeys.
    \item \textbf{VibeCraft} – \href{https://github.com/rayzhudev/vibecraft}{https://github.com/rayzhudev/vibecraft} \\
      Infinite pannable/zoomable canvas, folders/terminals/agents as draggable entities, git worktree tools.
    \item \textbf{GastownRTS} – \href{https://github.com/NJHFinanceHub/GastownRTS}{https://github.com/NJHFinanceHub/GastownRTS} \\
      Classic RTS HUD, minimap, resource bars, command cards, specialized agent role classes.
    \item \textbf{openage Engine} – \href{https://github.com/SFTtech/openage}{https://github.com/SFTtech/openage} + your fork.
\end{itemize}

\section{Core Concept: Creative Replacements}

We replace every AoE2 game element with AI-coding equivalents while preserving full RTS muscle memory:

\begin{longtable}{ll}
\toprule
\textbf{AoE2 Element} & \textbf{AgeOfAgents Replacement} \\
\midrule
Map / Terrain & Living architecture diagram of your monorepo (folders = terrain features, repos = regions) \\
Units (peasants, knights, etc.) & AI Agents (visual avatars with role icons; colors = status: idle/green, working/blue, error/red, attention/orange) \\
Buildings & Code modules, services, sub-repos, or folders (click to assign agents) \\
Resources (food, wood, gold, stone) & Compute (CPU/GPU), Tokens, Memory, Data \\
Tech Tree & Agent skill upgrades (new tools, RAG capabilities, Ollama model variants) \\
Fog of War & Unexplored code sections or unanalyzed dependencies \\
Campaign / Scenario Editor & High-level mission orchestrator (``Refactor entire protocol'') \\
Minimap & Overview of entire monorepo structure \\
Command Cards / Hotkeys & Spawn agent, Assign task, Review PR, Kill session, Cycle idle agents \\
Age of Empires II Definitive Edition graphics & Keep original assets for familiarity OR replace with custom sci-fi/AI-themed skins \\
\bottomrule
\end{longtable}

\section{Functional Requirements}

\subsection{1. Local-First Architecture}
\begin{itemize}
    \item 100\% offline capable. No cloud dependencies.
    \item Native Ollama integration (any local model via YAML config, exactly like Arcane Agents).
    \item Support for custom agent runtimes (Claude Code, OpenCode, Cursor CLI, or any bash/python wrapper).
    \item Persistent SQLite + tmux session storage (inherited and extended from Arcane).
\end{itemize}

\subsection{2. Monorepo \& Multi-Repo Discovery}
\begin{itemize}
    \item Auto-discovery engine (recursive scan, git worktrees, glob patterns).
    \item Point at root folder → entire protocol appears as procedural map.
    \item Hundreds of internal repos supported simultaneously.
\end{itemize}

\subsection{3. RTS Command Center UI (Merged Best Ideas)}
\begin{itemize}
    \item Infinite pannable/zoomable canvas (VibeCraft) + classic AoE2 camera controls.
    \item Classic RTS HUD: minimap, resource bars, quest log, command cards (GastownRTS).
    \item Units (agents) appear as draggable, selectable entities on the map.
    \item Live terminal streaming windows (Arcane Agents) that can be pinned or attached to units.
    \item Heat maps + collision detection for overlapping agent work (AgentCraft).
    \item Fog of war that lifts as agents explore code.
\end{itemize}

\subsection{4. Agent Spawning \& Orchestration}
\begin{itemize}
    \item Spawn agents via RTS controls or voice/text prompt.
    \item ``Campaign'' mode (AgentCraft): give high-level goal → agents decompose into quests and execute autonomously.
    \item Container isolation per agent (Docker or lightweight sandbox).
    \item Real-time status auto-detection from terminal output.
    \item Keyboard-first hotkeys for power users.
\end{itemize}

\subsection{5. Codebase-Aware Prompting \& Execution}
\begin{itemize}
    \item Attach agent to any building/folder → automatic context (files, git history, architecture diagram).
    \item Full prompting over any codebase (Ollama models receive RAG context from the monorepo).
    \item Agents can read, write, run tests, commit, and create PRs inside the game world.
\end{itemize}

\subsection{6. Review \& Collaboration}
\begin{itemize}
    \item Review bundles with visual evidence (screenshots, videos, diff heatmaps).
    \item Multiplayer ``Alliance Hall'' – shared battlefield for team + remote agents.
    \item Chat between human and agents; agents announce actions.
\end{itemize}

\subsection{7. Visual Architecture Representation}
\begin{itemize}
    \item The map \emph{is} your protocol architecture in real time.
    \item Buildings can be merged, split, or upgraded as you refactor.
    \item Resource nodes = critical files or dependencies.
\end{itemize}

\section{Non-Functional Requirements}
\begin{itemize}
    \item Performance: Handle 100+ simultaneous agents on consumer hardware.
    \item Extensibility: Full Python API (already in openage) + nyan modding system for new agent types.
    \item Security: All execution inside user-controlled containers; no external network unless explicitly allowed.
    \item Accessibility: Hotkey-first + optional voice commands.
    \item Cross-platform: Linux/macOS/Windows (via openage's existing support).
\end{itemize}

\section{Technical Architecture Overview}

\begin{itemize}
    \item \textbf{Core Engine}: Fork openage (C++ simulation + OpenGL/Qt6 UI).
    \item \textbf{New Backend Layer}: Node.js / Python bridge (reuse Arcane's tmux + runtime manager).
    \item \textbf{Ollama Integration}: YAML-defined runtimes that launch Ollama + custom agent scripts.
    \item \textbf{Modding}: Use nyan to define new unit types (agent roles), buildings (repo types), and techs (skills).
    \item \textbf{State Persistence}: SQLite for agent sessions + openage savegame format for entire battlefields.
\end{itemize}

\section{Implementation Roadmap (To-Do List for AI Agents)}

\subsection{Phase 0 – Fork \& Setup (1 week)}
\begin{enumerate}
    \item Fork \href{https://github.com/DylanCkawalec/openage}{your openage fork} and pull latest from upstream.
    \item Build and run the base engine locally (follow \texttt{doc/building.md}).
    \item Add a new ``AgeOfAgents'' modpack directory.
\end{enumerate}

\subsection{Phase 1 – Replace Game Logic (2-3 weeks)}
\begin{enumerate}
    \item Map monorepo root → procedural map generation.
    \item Replace unit AI with agent orchestration backend.
    \item Implement Arcane-style YAML runtime config for Ollama.
    \item Add live terminal streaming windows attached to units.
\end{enumerate}

\subsection{Phase 2 – UI \& RTS Features (3 weeks)}
\begin{enumerate}
    \item Infinite canvas + VibeCraft-style entity placement.
    \item Full GastownRTS HUD (minimap, resources, command cards).
    \item AgentCraft campaign/quest system.
    \item Status colors, heat maps, collision detection.
\end{enumerate}

\subsection{Phase 3 – Production Polish (2 weeks)}
\begin{enumerate}
    \item Container isolation + review bundles.
    \item Multiplayer Alliance Hall.
    \item Voice input + mobile PWA companion (optional but high value).
    \item Custom AI-themed graphics (replace AoE assets).
\end{enumerate}

\subsection{Phase 4 – Release \& Extensibility}
\begin{enumerate}
    \item Full documentation and example mod for new agent roles.
    \item Public GitHub release under GPLv3.
    \item Community modding guide using nyan + Python API.
\end{enumerate}

\section{Acceptance Criteria (For AI Coding Agents)}

The final system must:
\begin{itemize}
    \item Run completely locally with Ollama.
    \item Visually represent any monorepo as an AoE2-style battlefield.
    \item Allow spawning, commanding, and monitoring real coding agents via RTS controls.
    \item Preserve all listed ideas from the four source projects.
    \item Be extensible via Python and nyan.
    \item Feel like a production-ready masterpiece that makes managing 100+ agents feel like playing Age of Empires II.
\end{itemize}

\newpage
\section{Appendix A: Consolidated Feature DNA (All 4 Sources + openage)}

(Full list available in previous conversation extracts – every idea from AgentCraft video transcript, Arcane Agents README, VibeCraft canvas, GastownRTS HUD, and openage Python/nyan capabilities is to be implemented.)

This PRD is now complete and ready for implementation.

\textbf{Let's build the future of AI coding together.}

\end{document}